A supertree pipeline for summarizing phylogenetic and taxonomic information for millions of species

### 2017-01-26 doc/paper1-submission/Fig9.tex
Convert Fig9

\documentclass[fleqn,12pt,lineno,english]{wlpeerj}
\usepackage[T1]{fontenc}
%\usepackage[latin9]{inputenc}
%\usepackage{geometry}
\geometry{verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in}
%\usepackage{xcolor}
%\usepackage{verbatim}
\usepackage{float}
\usepackage{fancybox}
\usepackage{calc}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{natbib}
\sloppy
\makeatletter

%\def\fps@figure{ht}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}
\newcommand{\otl}{OTLv4}
\newcommand{\otlprop}{OTLv4$^\prime$}
\newcommand{\supportedby}{\texttt{supported\_by}}
\newcommand{\partialpathof}{\texttt{partial\_path\_of}}
\newcommand{\terminal}{\texttt{terminal}}
\newcommand{\conflictswith}{\texttt{conflicts\_with}}
\newcommand{\resolves}{\texttt{resolves}}
\newcommand{\resolvedby}{\texttt{resolved\_by}}

\floatstyle{ruled}
\newfloat{algorithm}{tbp}{loa}
\providecommand{\algorithmname}{Algorithm}
\floatname{algorithm}{\protect\algorithmname}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{algorithm,algpseudocode}
\usepackage{lmodern}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}
\definecolor{mylime}{RGB}{128,255,128}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

\usepackage{babel}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

\global\long\def\collections{\mathcal{C}}

\title{A supertree pipeline for summarizing phylogenetic and taxonomic information
for millions of species}

\author[1,2]{Benjamin D. Redelings}
\author[2,3,4]{Mark T. Holder}
\affil[1]{Department of Biology, Duke University, Durham NC, US}
\affil[2]{Department of Ecology and Evolutionary Biology, University of Kansas, Lawrence KS, US}
\affil[3]{Biodiversity Institute, University of Kansas, Lawrence KS, US}
\affil[4]{Heidelberg Institute for Theoretical Studies, Heidelberg, Germany}
\corrauthor[2,3,4]{Mark T. Holder}{mtholder@ku.edu}

% \keywords{Keyword1, Keyword2, Keyword3}

\begin{abstract}
We present a new supertree method that enables rapid construction of a
summary tree on the scale of millions of leaves.  This supertree
method summarizes a collection of input phylogenies and an input taxonomy.
We introduce formal goals and criteria for such a supertree to satisfy
in order to transparently and justifiably represent the input
trees. In addition to producing a supertree, our method computes
annotations that describe which grouping in the input trees support
and conflict with each group in the supertree.

\medskip

We compare our supertree construction method to a previously published
method by assessing their performance on input
trees used to construct the Open Tree of Life version 4, and find that
our method increases the number of displayed input splits from 35,518
to 39,639 and decreases the number of conflicting input splits from
2,760 to 1,357.  The new supertree method also improves on the previous
supertree construction method in that it produces no unsupported
branches and avoids unnecessary polytomies.

\medskip

This pipeline is currently used by the Open Tree of Life project to
produce all of the versions of project's ``synthetic tree'' starting
at version 5.This software pipeline is called
\emph{``propinquity.''}. It relies heavily on \emph{``otcetera''} - a
set of C++ tools to perform most of the steps of the pipeline. All of
the components are free software and are available on GitHub.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\section{Background}

The Open Tree of Life project seeks to build a platform for summarizing
what is known about phylogenetic relationships across all of Life
\citep{HinchliffEtAl2015}. One primary goal of the project is to
build a summary tree from a comprehensive taxonomic tree and a set
of published trees. The summary tree is intended to transparently
and justifiably represent phylogenetic information from these inputs.
The taxonomic tree is derived from the Open Tree Taxonomy (OTT hereafter;
publication in preparation). The phylogenetic inputs are published
trees that have been curated to align the tips to OTT and to identify
the correct rooting (see \citealt{McTavishEtAt2015} for further details
of the curation tools). Unlike OTT, these phylogenetic trees do not
include all leaf taxa. The inputs (taxonomy and phylogenetic trees)
and the output summary supertree are all rooted. Here we describe
the software pipeline (\emph{propinquity}) that summarizes and integrates
these smaller source trees and the taxonomy tree into a single supertree
and the noteworthy tools for manipulating and solving supertrees in
the \emph{otcetera} package.

\subsection{Goals}

Translating the goals of the Open Tree of Life's summary tree into
an explicit set of criteria is not trivial. The summary supertree
should represent the phylogenetic information from source trees in
a transparent and justifiable fashion. We would like to allow users
to correct errors in the supertree by improving the input information
rather than requiring modification to the supertree algorithm. The
pipeline was designed to create a tree which:
\begin{enumerate}
\item displays no unsupported groups,
\item defers to groupings from higher ranked trees in the case of conflict, 
\item contains no unnecessary polytomies, and
\item displays as many groupings from input trees as possible.
\end{enumerate}
These goals are described more fully below. In order to accomplish
transparency and justification, our pipeline also produces annotations
files with information about conflict and support.

\subsubsection{Goal 1: Each grouping is supported by at least one input}

We require that each edge in the supertree be supported by at least
one input tree edge. In addition to aiding interpretability, this
requirement keeps the supertree from arbitrarily representing information
that comes from none of the input trees. Of course, in a supertree
analysis, the full tree will imply some relationships for subsets
of the taxa that are not found in any input tree. So, the meaning
of ``supported by'' needs some clarification.

\paragraph*{Notation, terminology, and the definition of ``supported by''}

Cutting any edge $j$ of a rooted tree induces a bi-partition
$S(j)$ of the leaf taxa into two connected groups $S_{1}(j)$ and
$S_{2}(j)$. Such a bi-partition is called a rooted split and written
$S(j)=S_{1}(j)|\bullet S_{2}(j)$. Here $S_{1}(j)$ is called the ``cluster'' of $j$, and contains tip taxa on the side of edge $j$ that does not contain the
root. $S_{2}(j)$ contains the tip taxa on the side of edge $j$ that
contains the root. We refer to $S(j)$ as a \emph{rooted} split because
the right side of the split implicitly contains the root node, as indicated
by the $\bullet$.

For any two rooted splits $A=A_{1}|\bullet A_{2}$ and $B=B_{1}|\bullet B_{2}$,
we say that $A$ \emph{displays} $B$ if $B_{1}\subseteq A_{1}$ and
$B_{2}\subseteq A_{2}$. We also say that $A$ and $B$ \emph{conflict}
if none of $A_{1}\cap B_{1}$, $A_{1}\cap B_{2}$, and $B_{2}\cap A_{1}$
are empty.
Note that the intersection of the right sides of the splits is considered to
be non-empty even if $A_{2}\cap B_{2}$ is empty, because the right
sides implicitly contain the root. If $A$ and $B$ do not conflict, then we say that they are \emph{compatible}\footnote{It is possible to construct a rooted tree that displays both $A$
and $B$ if and only if $A$ and $B$ are compatible}. If $A$ displays $B$, then $A$ and $B$ must be compatible, since
$A_{1}\cap B_{2}$ and $B_{1}\cap A_{2}$ must be empty. We adopt
the shorthand of referring to an edge $k$ when we mean the split
$S(k)$. So, when we say that edge $j$ conflicts with edge $k$,
we mean that $S(j)$ conflicts with $S(k)$.

Let $\summaryTree$ denote a supertree, and $T_{i}$ denote the $i$th
input tree. We say that $\summaryTree$ displays an edge $j$ of $T_{i}$
if any edge of $\summaryTree$ displays $j$. We say that $\summaryTree$
conflicts with a edge $j$ of $T_{i}$ if any edge of $\summaryTree$
conflicts with $j$; otherwise $\summaryTree$ is compatible with
$j$. We say that edge $k$ of $\summaryTree$ is \emph{supported
by} edge $j$ of $T_{i}$ if $\summaryTree$ displays $j$, but $\summaryTree$
would not display $j$ if we contracted edge $k$. We also adopt the
shorthand of referring to edges by their tipward endpoint. By combining
this with the above shorthand, we could say that a node $j$ conflicts
or does not conflict with another node $k$.

The set of taxa that are mapped to the tips of the tree $T_{i}$ is
$\mathcal{L}(i)$. $\summaryTree(i)$ denotes the summary tree induced
by tip nodes that are mapped to taxa in $\mathcal{L}(i)$ and the
most recent common ancestor of those leaves, and any other node that
is an ancestor of some but not all of these leaves.

Note that stating that a node in the summary tree is supported by
an input does not imply that every descendant of that node must be
present in the input nor that every taxon that is not a descendant
must be excluded in order to display the node. Consider the problem
shown in Fig. \ref{fig:toyambig}; panels \ref{fig:toyambig}a
and \ref{fig:toyambig}b show two input trees. Because taxa A and
E do not occur together in either input, there is some uncertainty
about where to place them. By our terminology, either output shown
in Fig. \ref{fig:toyambig}c or Fig. \ref{fig:toyambig}d would be characterized
as a tree that displays all of the input groupings and which has no
unsupported groups. Clearly these criteria are insufficient to specify
a unique solution, and users of the output tree need to be aware that
it may be possible for some taxa to ``float'' to multiple positions.
In Fig. \ref{fig:toyambig}, taxon E floats to different positions
in \ref{fig:toyambig}c and \ref{fig:toyambig}d, whereas taxon
A does not.

\begin{figure}
\includegraphics[width=1.0\textwidth]{Fig1}

\caption{An example demonstrating that our definition of ``supported by''
does not imply entire composition of a grouping. (a) and (b) show
2 input trees and (c) and (d) depict trees that each display each
of the groupings in the input trees and which have no unsupported
nodes. The BUILD algorithm (section \ref{subsec:Subproblem-solution})
would choose tree (d) that floats taxon E closer to the root.}

\label{fig:toyambig}
\end{figure}

One of our aims in supertree construction is to minimize the amount
of information in the supertree that does not come from input trees.
We permit information that comes from combinations of input trees,
but not any single input tree. However, we seek to exclude information
that comes from none of the input trees. This motivates the criterion
of not having any unsupported edges, since these edges could be removed
without decreasing the support from any input tree.

\subsubsection{Goal 2: Tree ranking}

An appealing goal for the summarization would be to find the supertree
that displays the largest number of input tree edges. As discussed
in \citet[pages 92 and 131;][]{HusonRS2010} the maximum compatibility
problem is known to be $NP$-hard via a reduction to Max-Clique \citep{Karp1972}.
In addition to being computationally daunting, this formulation of
the supertree problem does not provide biologists who use the summarization
tool with an obvious avenue for fixing perceived problems with the
summary tree. For example, a grouping that a biologist expected may
not be present in the supertree, but it may not conflict with any
of the input groupings which are displayed. This can happen because
displaying both node $a$ from $T_{1}$ and node $b$ from tree $T_{2}$
in a summary tree may only be possible by displaying a grouping that
is present in no input tree. All other factors being equal, if this
implied grouping conflicts with input node $c$ in tree $T_{3}$,
then $c$ will not be displayed in the summary tree, but a biologist
will not necessarily know how to fix this problem. One solution is
to use a ranking of groupings. If an expert were quite confident in
the $c$ grouping, then she could assign that input node a high ranking.
A supertree that used ranks could then recover this grouping even
if its inclusion did not increase the total number of input nodes
that are displayed by the summary tree. Fig. \ref{fig:pairwisecompat}
shows an example of 3 input trees for which there is no pairwise incompatibility,
but no solution displays all of the input groups. Alternative rankings
of inputs can result in one of three summary trees shown in panels
d-f.

\begin{figure}
\includegraphics[width=1.0\textwidth]{Fig2}

\caption{An example of 3 input trees shown in (a), (b), and (c) which do not
conflict in a pairwise manner, but cannot be jointly displayed in
one tree. The 3 solution trees are shown in panels (d-f). Panel (d)
for ranking the tree in (c) lowest. Panel (e) shows the solution if
the tree in (b) has the lowest rank. Panel (f) shows the solution
if the tree in (a) is ranked lowest. Each of the solutions displays
2 of the 3 input groupings. }

\label{fig:pairwisecompat}
\end{figure}

Any approach to supertree construction must deal with the need to
adjudicate between conflicting input trees. We choose to deal with
conflict by ranking the input trees, and preferring to include edges
from higher-ranked trees. The merits of using tree ranking are questionable
because the system does not mediate conflicts based on the relative
amount of evidence for each alternative. However, it is a reasonable
starting point. It has the benefits of making it easy to see why some
groups are included or not (transparency), and it allows simpler and
cleaner algorithms.

Note that if some edge $c$ conflicts with a higher-ranked edge $b$,
then $c$ may still be included in the supertree. This can occur when
the higher ranked edge $b$ conflicts with a yet-higher ranked edge
$a$, and thus $b$ is not included. In that case, it will be possible
for $c$ to be represented in the summary tree. Thus, the fact that
the summary tree displays an input edge does not imply that none of
the higher ranked input trees conflict with that edge.

In order to produce a comprehensive supertree, we also require a rooted
taxonomy tree in addition to the ranked list of rooted input trees.
Unlike other input trees, the taxonomy tree is required to contain
all taxa, and thus has the maximal leaf set. We make the taxonomy
tree the lowest ranked tree. In our current formulation, the taxonomy
tree is also unique in that the taxonomy is the only source of taxonomic
names. Each node in the taxonomy tree corresponds to a named group.
Taxonomic groups may have the same name, but each node in the taxonomy
tree is identified by a unique number (its OTT ID). Taxonomic groups
are identified in the summary supertree by finding a branch (or ``node'')
that has exactly the same include|exclude relationship. The taxonomy
supertree can meaningfully possess degree-two nodes. Although these
nodes can be removed without affecting the relationships of the leaves,
they do represent nested taxonomic groups that contain exactly one
subgroup. The taxonomy is also used to determine which tips are terminal
taxa.

\subsubsection{Goal 3: Contain no unnecessary polytomies}

The supertree should be as resolved as possible - in other words,
it should have no unnecessary polytomies. Thus, for each input edge
that is \emph{not} included, we can point to a reason for non-inclusion
by showing that the input edge conflicts with some edge of the summary
tree. Note, that the requirement to not display unsupported groups
leads to some ``necessary'' polytomies. For example any resolution
of the polytomy shown in Fig. \ref{fig:pairwisecompat}e would
continue to display the same two input groups. However, the additional
grouping would be unsupported, because the unresolved tree already
displays both input groups. Thus, the 
unresolved tree would be preferred by our criterion.
However, collapsing either internal edge of the tree shown in Fig.
\ref{fig:pairwisecompat}d would result in a tree which displays
only one input grouping. This tree would contain an unnecessary polytomy,
because the polytomy would permit refinement to the depicted tree
which displays more input groupings.

\subsubsection{Goal 4: Display as many input nodes as feasible}

We also seek to construct a supertree that represents as many input
tree nodes as possible. Since non-included input tree nodes must conflict
with the supertree (or they would have been added), this criterion
is the same as minimizing the number of input nodes that conflict
with the supertree.

\subsubsection{Non-goal: phylogeny estimation}
Many researchers construct supertrees for the purpose of inferring the true tree from the input trees \citep{BinindaEmonds2007b,Davis2014}.  Such approaches treat input trees as data or as surrogates for data matrices \citep{gatesy2004critique}, and the supertree is seen as more accurate than the individual input trees because more input trees (and presumably more data) stand behind it.  For such methods, conflict resolution is a primary aim of the method.

Our supertree method operates in a different paradigm.  Phylogeny estimation is an explicit non-goal of our supertree algorithm.  We do not claim that the supertree is more accurate than the input trees it summarizes, and so conflict resolution is not a primary aim of the method. Most of our input trees have non-overlapping taxon sets, and so conflict resolution is not the primary problem.  Instead, we seek to merge and aggregate phylogenetic information on the largest possible scale.  Therefore, the primary aim of our supertree method is to \emph{summarize and represent} the input trees, even when they conflict with each other.  The annotations file constructed by our pipeline is essential for this task because it includes information about conflict with our supertree as well as support for it.

Our method must resolve conflicts in order to construct a single supertree.  However, the rank information used to resolve conflicts is an input to the method, not an output from the method.  We thus perform curation-based conflict resolution, not inference-based conflict resolution.

\subsubsection{Summary of goals}

These optimality criteria help to define what it means for the supertree
to represent the input trees, as well as justifying and explaining
why various features of the supertree exist. The pipeline described
below produces a supertree that satisfies the first three optimality
criteria and is a greedy approximation of a solution to the fourth
goal. It is not guaranteed to display as many input nodes as possible.
Even if the summary tree does accomplish goal 4, it is not necessarily
a \emph{unique} optimum. The pipeline takes a greedy approach to producing
a summary tree by attempting to add groupings from the trees in order
of the trees ranking. This can be viewed as a greedy solution to the
problem of finding the tree with the maximum sum of displayed groups'
weighted scores criterion (MSDGWS, described in the appendix A) where the
weights from the trees are so extreme that displaying one group from
a highly ranked tree is preferred to displaying all of the groupings
from lower ranked trees.  These large weights help to achieve our goal of allowing curators to fix problems in the tree.  While it is possible to view our pipeline as an optimization algorithm by assigning weights to the input trees in this way, we are not trying to weight the evidence in the input trees, and we do not actually ever calculate these weights in our pipeline.
Note that in this extreme form of weighting, we do not have to calculate
the score of a tree, we simply traverse the groupings by order of rank
and add a group if it is compatible with the previously added groups.

\subsubsection{Comparison to other supertree methods}

Matrix Representation Parsimony (MRP) is one of the
most widely used supertree approaches \citep{baum1992combining,BinindaEmonds2007b,Davis2014}.
MRP attempts to encode branches of input trees as characters in a data
matrix, and then find an optimal tree according to the maximum
parsimony criterion. Such an approach does not provide a clear and transparent
explanation for including particular edges in the supertree.
MRP can also infer edges that exist in none of the input trees, which 
violates our Goal \#1 \citep{gatesy2004critique}.

Another commonly used method is MinCut \citep{semple2000supertree} or
Modified MinCut \citep{page2002modified}.  Like our approach, MinCut
is based on the BUILD algorithm \citep{AhoSSU1981}.  However, MinCut deals
with conflicts by modifying the BUILD algorithm to resolve
incompatibilities by discarding edges that are present in the smallest 
number of input trees.  This approach thus violates our Goal \#2 of
resolving conflict via ranks that can be altered by a curator to
influence the output tree.

\section{Description of the supertree method}

\subsection{Preprocessing steps}

Propinquity was designed to function as a part of the Open Tree of
Life software architecture, so the first few steps of the pipeline
involve transforming artifacts from that project into a set of rooted
trees and a phylogenetic taxonomy. The phylesystem API \citep{McTavishEtAt2015}
of Open Tree allows users to curate published estimates of trees and
create ranked collections of these trees. Early steps in the propinquity
pipeline manipulate the phylogenetic input trees to improve their
usability and reliability. The first steps of the pipeline (see Fig.
\ref{fig:pipeline}) collect a list of trees to include (in the \texttt{phylo\_input}
subdirectory) and store copies of these files (in the \texttt{phylo\_snapshot}
subdirectory) to make it easier to replicate the operation (because
the collection of trees and the tree files change due to curation).
\begin{figure}
\begin{centering}
\includegraphics[height=0.70\paperheight]{Fig3.pdf}
\par\end{centering}
\caption{Organization of the \emph{propinquity} pipeline. Each colored pentagon
labels a program (blue for otcetera-based tools and red for python
scripts in the propinquity or peyotl repository) that performs the
important operations in each step; the number before the tool name
refers to the section in this paper that describes the operation.
The output of each step corresponds to a subdirectory of the propinquity
system which will hold the output artifacts for the step. Ovals are
resources that are required (OTT and Open Tree's phylesystem repository).
White pentagons are user-controlled inputs.}
\label{fig:pipeline}
\end{figure}

\subsubsection{Pruning questionable taxa from the taxonomy}

OTT is a hierarchy of taxonomic names that implies a phylogenetic
taxonomy. An OTT ID has a position in the hierarchy, a taxonomic name,
and set of references to the same name in different taxonomies. In
addition, the ID may also be associated with a set of flags that can
indicate that the taxon may be questionable. These flags can either
encode information taken from an input taxonomy (for example, taxa
the NCBI refers to as ``unplaced'' are assigned an ``unplaced''
flag) or can arise because of some form of conflict during taxonomy
construction (for example, if two taxonomies disagree on the name
for a taxon, then the taxon can be merged and the name will be retained
without any descendants; this name will have an OTT ID, but will be
flagged as ``barren''). Propinquity prunes the OTT down to a more
reliable taxonomy by pruning off parts of the tree that are flagged
with suspicious flags. The set of flags that lead to a subtree of
the taxonomy being pruned is under the control of the user (the set
of flags used by the Open Tree of Life project can be found in the
\texttt{config.opentree.synth} file in the propinquity repository).
For the purpose of the rest of the pipeline, an OTT ID that has been
pruned from the taxonomy will be treated in the same way as invalid
OTT ID. The output of this step is stored in propinquity's \texttt{cleaned\_ott}
subdirectory; this operation only needs to be performed when the OTT
or the pruning flags change.

\subsubsection{Pruning problematically mapped tips from input phylogenetic trees}

Frequently, phylogenetic estimates are rooted using the outgroup criterion,
which is an assumption about the monophyly of the ingroup taxa. Because
the rooting of the branches in the outgroup portion of the tree is
often uncertain, data curators can identify the ingroup node of the
tree; propinquity uses this annotation to prune off the outgroup taxa.

Frequently, not all tips in a phylogenetic input will have been mapped
to a taxon in the current version OTT. Unmapped leaves are pruned
from each phylogenetic input. In some cases, the OTT has changed and
a taxon has been unambiguously mapped to another taxon. This can occur
when multiple species in one version of the taxonomy are ``lumped''
into a single taxon in a subsequent version. OTT maintains a set of
``forwarding'' statements about IDs that have been removed but can
be mapped to an existing taxon; propinquity uses these statements
to update the OTU mapping of input trees.

Finally some leaves are mapped to taxa that occur more than once in
the tree, or taxa that have ancestors represented as tips of the tree.
In these cases, leaves are pruned to assure that tips are mapped to
unique taxa that are not nested. In the case of nested taxa, the tip
mapped to the higher level taxon is pruned, and one of the lower level
tips is retained. In the case of duplicate occurrences of an OTT taxon,
propinquity checks to see if a data curator has selected one of the
taxa to be the exemplar for the taxon. If this selection has not been
made, then the node with the lexicographically lowest ID is chosen
to exemplify the taxon. This choice is arbitrary, but repeatable.
The pruned phylogenetic inputs are stored in a \texttt{cleaned\_phylo}
subdirectory of propinquity.

\subsubsection{Exemplifying tips mapped to higher taxa}

Many input trees have tips that are not terminal taxa, but higher-order
taxonomic groups. It is not clear how to interpret a tip in a phylogenetic
estimate that is labeled with the name of a higher taxon. Several
scenarios can lead to these cases: the data for the tip could have
been created by merging a chimeric set of character scores from constituent
taxa; the species sampled may not have been identified to the lowest
taxonomic rank; or the researcher may simply have used a higher taxonomic
name because he/she assumed that the taxon is monophyletic and the
higher level name would be more recognizable. Rather than allowing
the ambiguity about interpretation of the higher-taxon mapped tips
to propagate throughout the entire pipeline, we transform the input
trees by replacing higher taxa at tips with a set of terminal-taxon
exemplars for each taxon. One approach would be to simply determine
all descendant terminal taxa and attach them as children of the problematic
tip. However, this would create a clade rather than a tip; subsequent
steps in the supertree would interpret the clade as a claim of monophyly
for the taxon. The input tree may not have tested monophyly of the
clade, so this interpretation is unwarranted. We avoid it by attaching
exemplar taxa as child nodes of the higher taxonomic tip but then
collapsing the edge between the former tip node and
its parent. Thus, if $A$ is a non-terminal taxon containing terminal
descendants $a_{1}$ and $a_{2}$ and $B$ is a non-terminal taxon
containing terminal descendants $b_{1}$ and $b_{2}$ we would replace
the subtree $((A,B),c)$ with $((a_{1},a_{2},b_{1},b_{2}),c)$ instead
of the subtree $(((a_{1},a_{2})A,(b_{1},b_{2})B),c)$.

If a taxon is only present in the taxonomy (not in any of the input
trees), then it can be pruned from the taxonomy for the construction
of the supertree and then grafted back on to the summary tree later.
Performing this pruning reduces the size of the supertree problem,
reducing the running time of the pipeline. Similarly, when we expand
a higher taxon in the exemplification step, we can omit members of
the taxon if they do not occur in any of the phylogenetic inputs.
If there are no members of the higher taxon sampled in any other input
tree, then we arbitrarily choose one terminal taxon to represent the
higher taxon. During the exemplification step, a tool from otcetera
(\texttt{otc-nonterminals-to-exemplars}) reads the taxonomy and all
of the ``cleaned'' phylogenetic estimates from the previous step.
Reading all of the inputs is necessary to assure that each higher
taxon is replaced with the same set of exemplars regardless of which
tree the higher taxon occurs in, and that the exemplars for a higher
taxon is the union on the set of descendant terminal taxa that have
been sampled in a phylogenetic input. For example, the trees in Fig.
\ref{fig:cleaned-phylo-example} would exemplified as shown in Fig.
\ref{fig:exemplify-example}.

We prune the taxonomy by removing tips that are not present in any
input tree to produce the pruned taxonomy $\mathbb{T}_{p}$. The tips
pruned in this step will be grafted back onto the skeleton of the
summary tree in a subsequent step. A terminal taxon that is represented
only in the taxonomy can be pruned and then regrafted onto the solution
without affecting which nodes are displayed by the final summary tree.
Thus, this procedure does not impede our ability to find a good summary
tree. Removing these tips produces a smaller input to the rest of
the pipeline, which reduces running times. After producing the set
of ``exemplified'' phylogenetic inputs, this tool exports a pruned
down version of the taxonomy that only contains tips that are present
in at least one phylogenetic input.

\begin{figure}
\includegraphics[width=1.0\textwidth]{Fig4}

\caption{Input trees (a-b) and taxonomy tree (c).}
\label{fig:cleaned-phylo-example}
\end{figure}

\begin{figure}
\includegraphics[width=1.0\textwidth]{Fig5}

\caption{Exemplified input trees (a-b) and taxonomy tree (c) from Fig. \ref{fig:cleaned-phylo-example}.
Taxon E in (a) is exemplified by E1. Pruned taxa are E2, F2, and D. The
taxa E and F are retained as monotypic taxa in the pruned taxonomy
$\protect\prunedTaxonomy$, but are not shown in panel (c). The red
edge in the pruned taxonomy tree is an uncontested higher taxon in
the exemplified taxonomy (as explained in section \ref{subsec:Subproblem-decomposition})}
\label{fig:exemplify-example}
\end{figure}

\subsection{Summary tree construction}

After the preprocessing steps, the inputs have been converted to a
set of rooted phylogenetic estimates in which each leaf is mapped
to a terminal taxon in the exemplified taxonomic tree. The goal of
the remainder of the pipeline is to construct a tree that maximizes
the sum of displayed groups' weighted scores (MSDGWS) criterion. This is accomplished
in four steps: (1) dividing the full problem into subproblems based
on uncontested taxa, (2) constructing a summary solution for each
subproblem by greedily creating a maximally-sized list of groupings
that can all be displayed simultaneously; (3) grafting the subproblem
solutions into a single supertree; and (4) grafting (or ``unpruning'')
the taxonomy-only taxa onto the solution to produce a complete summary
tree.

\subsubsection{Subproblem decomposition\label{subsec:Subproblem-decomposition}}

For the sake of efficiency, propinquity uses a divide-and-conquer
approach to construct the supertree. Subproblems are identified by
searching through the taxonomy tree to find any taxa that are not contested
by any single input tree. Here we say that input tree $T_{i}$ contests
taxon $x$ in the pruned taxonomy, if $x$ is not monophyletic in
any resolution of tree $T_{i}$. Thus, polytomies in an input tree
are treated as soft polytomies, and a taxon is not contested merely
because it is not displayed by an input tree.

This operation is performed by the \texttt{otc-uncontested-decompose}
tool in otcetera; see appendix \ref{sec:Decompose-algorithm} for
a description of the algorithm. The output is a series of subproblems,
each of which corresponds to a slice of the taxonomy and corresponding
slices through each relevant input tree. Each uncontested non-terminal
and non-root taxon will show up in two subproblems: it will be the
root of its own subproblem and it will be tip in the subproblem that
covers the next slice deeper in the tree. The red edge in Fig. \ref{fig:exemplify-example}c
highlights the taxa that are not contested by the input shown in Fig. \ref{fig:exemplify-example};
Fig. \ref{fig:subproblems-example} shows the subproblems that would
be emitted as a result of this set of inputs. The supertree operation
of \citet{HinchliffEtAl2015} also used this otcetera-based decomposition
step.

Note that decomposition into uncontested groups does not necessarily
allow us to find the tree that maximizes the MSDGWS score. For example,
see Fig. \ref{fig:decompose-worsens}; that example is a variant
of the situation shown in Fig. \ref{fig:pairwisecompat}. In this
case the groupings from each of the phylogenetic estimates, shown
in panels \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b,
could be displayed. That solution is shown in panel \ref{fig:decompose-worsens}d,
it displays two of the three input splits, but is optimal because
no solution displays all three input groupings and the depicted solution
displays the two highest ranked groupings. However, neither of the
trees shown in panels \ref{fig:decompose-worsens}a or \ref{fig:decompose-worsens}b
contest the taxon $B$ shown in the taxonomy panel \ref{fig:decompose-worsens}c.
Thus, when using our decomposition, the branches leading to taxa $B1$
and $B2$ in the input phylogenetic trees would be sliced during the
decomposition, and relabeled to refer to taxon $B$. This taxonomically-informed
interpretation of the inputs views the two phylogenetic inputs as
in conflict; so the solution returned by propinquity would defer to
the higher ranked tree. The tree shown in panel \ref{fig:decompose-worsens}e
would be returned. This example arises from the fact that the trees
in \ref{fig:decompose-worsens}a and \ref{fig:decompose-worsens}b
jointly contest taxon $B$, but neither contests taxon $B$ when the
trees are considered in isolation.

Despite the fact that the use of \texttt{otc-uncontested-decompose}
can worsen the final score of the summary tree, we use this approach
in propinquity because it makes the construction of the tree faster
and it is easy for users to correct issues caused by incorrect taxa
being constrained to be monophyletic. By adding a tree (even a low-ranked
tree) that contests a taxon to the corpus of input trees , then the
next synthetic tree will no longer consider the taxon to be uncontested.
Thus the procedure encourages curation of more phylogenetic inputs
as a means of improving the summary tree.

\begin{figure}
\includegraphics[width=1.0\textwidth]{Fig6}

\caption{Subproblems \emph{ABCD} and \emph{root} generated from the exemplified trees shown in Fig. \ref{fig:exemplify-example}.
A trivial statement from the first tree that a taxon labelled $ABCD$
is sister to $E$ has been omitted, because trees with only 2 leaves
do not contain phylogenetic information.
Similarly, the trivial problems that list E1 and F1 as the only daughters
of taxa E and F have been omitted.}
\label{fig:subproblems-example}
\end{figure}
\begin{figure}
\includegraphics[width=1.0\textwidth]{Fig7}

\caption{An example with three input trees: the highest ranked phylogenetic
input panel (a), the second ranked phylogenetic input (b), and the
taxonomy in panel (c). The summary tree in panel (d) has the highest
possible score, but the summary shown in panel (e) would be returned
from the pipeline that uses uncontested taxon decomposition.}
\label{fig:decompose-worsens}
\end{figure}

\subsubsection{Subproblem solution\label{subsec:Subproblem-solution}}

When solving sub-problems, we sequentially incorporate splits from
trees in order of ranking, retaining splits that are compatible with
the current set of splits (Fig. \ref{alg:ConsistentSplitsFromRankedList}).
The order of splits from the same tree is not specified by this approach,
and we incorporate splits using one of the possible post-order traversals
of the tree. We make use of the BUILD algorithm \citep{AhoSSU1981}
to assess compatibility. This strategy avoids unnecessary polytomies,
since splits of later input trees are only rejected from the summary
supertree if they conflict with higher-priority splits. Finally, we
use the BUILD algorithm to construct a supertree displaying all of
the splits in the set of compatible splits. Using the BUILD algorithm
to construct the subproblem summary tree satisfies criterion 3, because
trees from the BUILD algorithm do not contain unsupported branches.
\begin{figure}
\includegraphics[width=0.9\textwidth]{fig8}
\caption{\label{alg:ConsistentSplitsFromRankedList}Algorithm \texttt{ConsistentSplitsFromRankedList}}
\end{figure}

The BUILD algorithm as originally stated by \citet{AhoSSU1981} applies
to a collection of rooted triplets. Instead of decomposing each input
split into a collection of rooted triplets, we instead modify the
BUILD algorithm to apply directly to larger rooted splits. The modified
BUILD algorithm constructs a tree compatible with a collection of
rooted splits, and returns failure if such a tree does not exist.
This modified algorithm recovers the original BUILD algorithm if only
rooted triplets are supplied as input. When larger splits are supplied
as input, the results are the same as if each was was decomposed into
all implied triplets. The modified build algorithm has order $O(N^{2}+N^{2}E+NL)$
where $N$ is the number of splits passed in, $E$ is the average
size of the exclude group, and $L$ is the total number of leaves.
This simplifies to $O(N^{2})$ if all splits are triplets. In this
approach splits are either entirely retained or entirely discarded
- consistent rooted triplets from conflicting splits are not retained.
However, when unpruning taxonomy-only taxa (see below), we make an
attempt to break ties in a way that preserves some partial information
from conflicting splits by attaching taxa from conflicting splits
at their common ancestor. Fig. \ref{fig:subproblem-solutions.}
shows the solutions that would be obtained by applying our modified
version of the BUILD algorithm to the subproblems shown in Fig. \ref{fig:subproblems-example}.

\begin{figure}
\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Fig9a}

@@

\caption{Solutions to subproblem \emph{ABCD} and subproblem \emph{root} depicted in Fig. \ref{fig:subproblems-example}.}
\label{fig:subproblem-solutions.}
\end{figure}

\subsubsection{Solution grafting}

>>>>

\documentclass[12pt,preview]{standalone}
\usepackage[labelfont={bf,sf}]{caption}
\usepackage[verbose,tmargin=1in,bmargin=1in,lmargin=1in,rmargin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{lmodern}
\usepackage{subfig}
\usepackage{amssymb}
%\usepackage{subcaption}
\global\long\def\taxonomy{\mbox{\ensuremath{\mathbb{T}}}}

@@

\global\long\def\collections{\mathcal{C}}
\begin{document}
\begin{figure}
\hfill{}\subfloat[Solution to subproblem \emph{ABCD}]{\includegraphics[scale=0.55]{Fig9a}

@@

\end{figure}
\end{document}
